Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-04-06 paper.tex
Update table 1

\usepackage{authblk}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{makecell}
\usepackage{url}
% JK: turning this off for the moment as I keep clicking through on links
% to the bibliography while reading the text and it's intensely annoying.
% Can reinstate when we're ready to preprint
% \usepackage[hidelinks]{hyperref}

@@

\begin{table}
\centering
\begin{tabular}{l|l|l} \hline \textbf{Property} &
\textbf{Wide ARG} & \textbf{Long ARG} \\ \hline Last sampling date & 2021-06-30
& 2022-06-30 \\ Local trees & 1,496 & 958 \\ \hline
\multicolumn{3}{l}{\textit{Nodes}} \\ All nodes (sample and non-sample) &
1,453,347 & 783,231 \\ Sample nodes & 1,265,685 (87\%) & 657,239 (84\%) \\ Tree
adjustment nodes (excluding recombination nodes) & 80,830 (6.34\%) & 72,107
(10.97\%) \\ Recombination nodes & 4,123 (0.28\%) & 2,078 (0.27\%) \\ \hline
\multicolumn{3}{l}{\textit{Types of recombination nodes}} \\ Singletons (i.e.
no descendant besides the sequence causing insertion of the node) & 2,601
(63\%) & 1,315 (63\%) \\ Two-parent recombinants & 3,521 (85\%) & 1,869 (90\%)
\\ HMM-consistent recombinants & XX (XX\%) & XX (XX\%) \\ \hline
\multicolumn{3}{l}{\textit{Mutations}} \\ Total mutations & 1,213,193 &
1,062,072 \\ Mutations on a terminal edge & 758,903 (62.55\%) & 767,111
(72.23\%) \\ Recurrent mutations & 74,719 (6.16\%) & 50,099 (4.72\%) \\
Reversions & 72,617 (5.99\%) & 48,226 (4.54\%) \\ Immediate reversions (one
intervening node only) & 3,541 (0.29\%) & 1,347 (0.13\%) \\ \hline
\multicolumn{3}{l}{\textit{Storage and loading time}} \\ tszipped file
(including sample metadata, imputed Pango lineage assignments, and results of
HMM runs for recombination nodes) & ~57 Mb & ~36 Mb \\ Bytes to encode one
sample & 8.29 & 10.83 \\ Load time (seconds) & 0.9 & 0.5 \\ \hline
\end{tabular}
\caption{\label{tab:args}Summary of the inferred ARGs.
}
\end{table}

@@

The wide ARG (with over one million sample genomes) was built using 128 threads
on a machine with 512 GB RAM (2x AMD EPYC 7502 @ 2.5GHz). Inference ran for 17
days to attach genomes sampled up to June 30, 2021. The long ARG (with ~700
thousand sample genomes) was built using 40 threads on a machine with 256 GB
RAM (2x Intel(R) Xeon(R) CPU E5-2680 v2 @ 2.80GHz). Inference ran for 23 days
to attach genomes sampled up to June 30, 2022. Most of the runtime was spent on
running the HMM.

>>>>

\usepackage{booktabs}
\usepackage{authblk}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{makecell}
\usepackage{url}
% JK: turning this off for the moment as I keep clicking through on links
% to the bibliography while reading the text and it's intensely annoying.
% Can reinstate when we're ready to preprint
% \usepackage[hidelinks]{hyperref}

@@

\begin{table}
\begin{center}
    \begin{tabular}{llrlrl}\toprule
        &  & \multicolumn{2}{l}{Wide ARG} & \multicolumn{2}{l}{Long ARG} \\
    \cmidrule{3-6}
    \multicolumn{2}{l}{Sample filtering}
             & \multicolumn{2}{l}{collection $\leq$ 2021-06-30}
                 & \multicolumn{2}{l}{collection $\leq$ 2022-06-30} \\
           & & \multicolumn{2}{l}{max-delay=30} & \multicolumn{2}{l}{max-delay=30} \\
        &  & &                               & \multicolumn{2}{l}{max-daily=1000} \\
    \midrule

    Nodes & & 1,453,347 & & 783,231  & \\
    & Samples & 1,265,685 & (87\%) & 657,239 & (84\%) \\
    & Local (UPGMA) & 103K & (7.1\%) & 52K & (6.6\%)  \\
    & Tree adjustment & 80,830 & (5.6\%) & 72,107 & (9.2\%) \\
    & Recombination & 4,123 & (0.33\%) & 2,078 & (0.27\%) \\
    \cmidrule{2-6}
    % TODO add more information on mutations
    Mutations & & 1.21M &  & 1.06M & \\
    & On one sample & 759K & (63\%) & 767K & (72\%) \\
    & Reversions & 72K & (5.98\%) & 48K & (4.54\%) \\
    & per node & 0.83  &   & 1.36 \\
    \cmidrule{2-6}
    Recombination events & \\
    & Total & 4,123 & & 2,078 & \\
    & Singletons & 2,601  & (63\%) & 1,315 & (63\%) \\
    % clearer for many than "local trees?"
    & Unique breakpoints & 1,496 &  & 958 & \\
    \cmidrule{2-6}
    \multicolumn{2}{l}{Size (inc metadata)} & 57MB & & 36MB&  \\
    \multicolumn{2}{l}{Bytes/sample (exc metadata)}  & 8.29  & & 10.83 \\
    Load time & & 0.9s & & 0.5s & \\
    \bottomrule
\end{tabular}
\end{center}

\caption{\label{tab:args}Summary of the inferred ARGs.
}
\end{table}

% \centering
% \begin{tabular}{l|l|l} \hline \textbf{Property} &
% \textbf{Wide ARG} & \textbf{Long ARG} \\ \hline Last sampling date & 2021-06-30
% & 2022-06-30 \\ Local trees & 1,496 & 958 \\ \hline
% \multicolumn{3}{l}{\textit{Nodes}} \\ All nodes (sample and non-sample) &
% 1,453,347 & 783,231 \\ Sample nodes & 1,265,685 (87\%) & 657,239 (84\%) \\ Tree
% adjustment nodes (excluding recombination nodes) & 80,830 (6.34\%) & 72,107
% (10.97\%) \\ Recombination nodes & 4,123 (0.28\%) & 2,078 (0.27\%) \\ \hline
% \multicolumn{3}{l}{\textit{Types of recombination nodes}} \\ Singletons  & 2,601
% (63\%) & 1,315 (63\%) \\ Two-parent recombinants & 3,521 (85\%) & 1,869 (90\%)
% \\ HMM-consistent recombinants & XX (XX\%) & XX (XX\%) \\ \hline
% \multicolumn{3}{l}{\textit{Mutations}} \\ Total mutations & 1,213,193 &
% 1,062,072 \\ Mutations on a terminal edge & 758,903 (62.55\%) & 767,111
% (72.23\%) \\ Recurrent mutations & 74,719 (6.16\%) & 50,099 (4.72\%) \\
% Reversions & 72,617 (5.99\%) & 48,226 (4.54\%) \\ Immediate reversions (one
% intervening node only) & 3,541 (0.29\%) & 1,347 (0.13\%) \\ \hline
% \multicolumn{3}{l}{\textit{Storage and loading time}} \\ tszipped file
%  & ~57 Mb & ~36 Mb \\ Bytes to encode one
% sample & 8.29 & 10.83 \\ Load time (seconds) & 0.9 & 0.5 \\ \hline
% \end{tabular}

@@

Inferring the wide ARG required 17 days of computation time on a machine
with 128 threads and 512 GB RAM (2x AMD EPYC 7502 @ 2.5GHz).
The long ARG required 23 days on a (much older) machine
with 40 threads and 256 GB RAM (2x Intel(R) Xeon(R) CPU E5-2680 v2 @ 2.80GHz).
The majority of the time is spent on running the HMM to find the copying
path for each sequence, and the process is therefore highly amenable to
distributing across multiple machines.
